---
title: "Analyze surveychat transcripts with R"
output: html_document
---

## Goal

This short example reads a Qualtrics CSV file and turns the surveychat JSON
transcript column into a table with one row per chat message.

Only edit the two lines in the first code block: the CSV file name and the
column that contains the JSON transcript.

```{r setup}
library(tidyverse)
library(jsonlite)
```

## 1. Read the CSV

```{r read-csv}
# Edit these two lines.
csv_file <- "qualtrics_export.csv"
json_column <- "chat_transcript"

responses <- read_csv(csv_file, show_col_types = FALSE)

if (!json_column %in% names(responses)) {
  stop(
    "Column not found. Check json_column. Available columns are: ",
    str_c(names(responses), collapse = ", ")
  )
}

transcripts <- responses %>%
  mutate(
    response_number = row_number(),
    transcript_json = as.character(.data[[json_column]])
  ) %>%
  filter(!is.na(transcript_json), str_trim(transcript_json) != "")
```

## 2. Parse the JSON transcripts

```{r parse-json}
# This is just a small check. Rows listed here are not valid JSON.
is_valid_json <- function(transcript_json) {
  tryCatch(
    {
      fromJSON(transcript_json)
      TRUE
    },
    error = function(e) FALSE
  )
}

malformed_json <- transcripts %>%
  filter(!map_lgl(transcript_json, is_valid_json))

parse_messages <- function(transcript_json, response_number) {
  tryCatch(
    {
      parsed <- fromJSON(transcript_json)

      if (!"messages" %in% names(parsed) || length(parsed$messages) == 0) {
        return(tibble())
      }

      as_tibble(parsed$messages) %>%
        mutate(
          response_number = response_number,
          turn_number = row_number(),
          .before = 1
        )
    },
    error = function(e) tibble()
  )
}

messages <- map2_dfr(
  transcripts$transcript_json,
  transcripts$response_number,
  parse_messages
)

if (nrow(messages) == 0) {
  stop("No messages were parsed. Check csv_file and json_column.")
}

messages <- messages %>%
  mutate(
    role = as.character(role),
    content = replace_na(as.character(content), ""),
    timestamp = parse_datetime(timestamp),
    word_count = if_else(
      content == "",
      0L,
      str_count(str_squish(content), "\\S+")
    )
  )

messages
```

## 3. Simple conversation metrics

```{r conversation-summary}
conversation_summary <- messages %>%
  group_by(response_number) %>%
  summarise(
    n_turns = n(),
    participant_turns = sum(role == "participant", na.rm = TRUE),
    assistant_turns = sum(role == "assistant", na.rm = TRUE),
    participant_words = sum(
      if_else(role == "participant", word_count, 0L),
      na.rm = TRUE
    ),
    assistant_words = sum(
      if_else(role == "assistant", word_count, 0L),
      na.rm = TRUE
    ),
    .groups = "drop"
  )

conversation_summary
```

```{r role-summary}
role_summary <- messages %>%
  group_by(role) %>%
  summarise(
    n_messages = n(),
    mean_words = mean(word_count, na.rm = TRUE),
    .groups = "drop"
  )

role_summary
```

## 4. Simple text example

This combines all participant messages from each response. You can use this
table for hand coding, keyword checks, or later text analysis.

```{r participant-text}
participant_text <- messages %>%
  filter(role == "participant") %>%
  group_by(response_number) %>%
  summarise(
    participant_text = str_c(content, collapse = " "),
    .groups = "drop"
  )

participant_text
```

```{r keyword-example}
# Change this word to match your research question.
keyword <- "trust"

keyword_summary <- participant_text %>%
  mutate(
    mentions_keyword = str_detect(
      str_to_lower(participant_text),
      str_to_lower(keyword)
    )
  ) %>%
  summarise(
    n_responses = n(),
    responses_mentioning_keyword = sum(mentions_keyword),
    percent_mentioning_keyword = mean(mentions_keyword) * 100
  )

keyword_summary
```
